---
title: "Investigating the Impact of Local Meteorological Conditions on Daily Air Pollutant Levels in Houston Using The Generalized Additive Model"
author: "Joseph Ephraim-Addison"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: powerpoint_presentation
subtitle: 
---


                                  ### JUST IGNORE ###

```{r eval=FALSE, message=FALSE, comment="", include=FALSE}


data1 <- read.csv("C:/Users/Ephra/OneDrive/Documents/New folder/PM2.5/ad_viz_plotval_data.csv")

data2 <- read.csv("C:/Users/Ephra/OneDrive/Documents/New folder/PM2.5/ad_viz_plotval_data (1).csv")

data3 <- read.csv("C:/Users/Ephra/OneDrive/Documents/New folder/PM2.5/ad_viz_plotval_data (2).csv")

data4 <- read.csv("C:/Users/Ephra/OneDrive/Documents/New folder/PM2.5/ad_viz_plotval_data (3).csv")

data5 <- read.csv("C:/Users/Ephra/OneDrive/Documents/New folder/PM2.5/ad_viz_plotval_data (4).csv")

data6 <- read.csv("C:/Users/Ephra/OneDrive/Documents/New folder/PM2.5/ad_viz_plotval_data (5).csv")

data7 <- read.csv("C:/Users/Ephra/OneDrive/Documents/New folder/PM2.5/ad_viz_plotval_data (6).csv")

data8 <- read.csv("C:/Users/Ephra/OneDrive/Documents/New folder/PM2.5/ad_viz_plotval_data (7).csv")

data9 <- read.csv("C:/Users/Ephra/OneDrive/Documents/New folder/PM2.5/ad_viz_plotval_data (8).csv")

data10 <- read.csv("C:/Users/Ephra/OneDrive/Documents/New folder/PM2.5/ad_viz_plotval_data (9).csv")

data11 <- read.csv("C:/Users/Ephra/OneDrive/Documents/New folder/PM2.5/ad_viz_plotval_data (10).csv")

data13 <- read.csv("C:/Users/Ephra/OneDrive/Documents/New folder/PM2.5/ad_viz_plotval_data (12).csv")

data14 <- read.csv("C:/Users/Ephra/OneDrive/Documents/New folder/PM2.5/ad_viz_plotval_data (13).csv")

data15 <- read.csv("C:/Users/Ephra/OneDrive/Documents/New folder/PM2.5/ad_viz_plotval_data (14).csv")

data16 <- read.csv("C:/Users/Ephra/OneDrive/Documents/New folder/PM2.5/ad_viz_plotval_data (15).csv")

data <- rbind(data1,data2, data3, data4, data5, data6, data7, data8, data9, data10, data11, data13, data14, data15, data16)

dim(data)

#data <- data %>% group_by(Date)

#write_xlsx(data, "data.xlsx")


```

                        ### AIR QUALITY DATA ANALYSIS ###
                        
```{r message=FALSE, comment="", include=FALSE, results='hold'}
library(readxl)
library(writexl)
library(dplyr)
library(lubridate)
library(tidyr)


# List of file paths
file.paths <- list.files(path = "C:/Users/Ephra/OneDrive/Desktop/New folder/PM2.5", pattern = "ad_viz_plotval_data.*\\.csv", full.names = TRUE)


# Read and merge all data files
merged_data <- lapply(file.paths, read.csv) %>%
  bind_rows()


Airdata <- merged_data %>% rename(PM2.5.Level = Daily.Mean.PM2.5.Concentration, Site.Name = Local.Site.Name, Latitude = Site.Latitude, Longitude = Site.Longitude)

str(Airdata)


#Airdata %>% View()



```



```{r message=FALSE, comment="", include=FALSE, results='hold'}

#unique(Airdata$Date)

Airdata$Year <- year(mdy(Airdata$Date))
Airdata$Dow <- day(mdy(Airdata$Date))
Airdata$Month <- month(mdy(Airdata$Date))


#write_xlsx(Airdata, "Air.xlsx")

```


```{r message=FALSE, comment="", include=FALSE, results='hold'}
sum(Airdata$Year == 2009)

get_season <- function(Month) {
  if (Month %in% c(12, 1, 2))  {
    return("Winter")
  } else if (Month %in% c(3, 4, 5)) {
    return("Spring")
  } else if (Month %in% c(6, 7, 8)) {
    return("Summer")
  } else {
    return("Fall")
  }
}

Airdata$Season <- sapply(Airdata$Month, get_season)

  
str(Airdata)


```


```{r message=FALSE, comment="", include=FALSE, results='hold'}

Air <- Airdata %>% mutate(Site.ID = as.character(Site.ID)) %>% 
  select(1, 3, 5, 8, 13, 14, 21, 22, 23, 24, 25, 26 )


Air <- Air %>% filter(Site.ID %in% c("481671034", "482010024", "482011034", "482011035", "482011039", "482011050", "483390078", "482011042", "482010416"), Method.Description %in% c("PM2.5 SCC w/Correction Factor"))


#colSums(is.na(Air))

#summary(Air$Site.Name == "Baytown")

#length(unique(Air$Site.Name))
#length(unique(merged_data$Site.ID))

#str(Air)


```

```{r message=FALSE, comment="", include=FALSE}
PM2.5 <- Air %>% filter(Year %in% c(2009, 2010, 2011, 2012)) %>% select(1, 2, 4, 8, 7, 10, 11, 9, 12, 3)

summary(PM2.5$Site.ID == "482010416")

#colnames(PM2.5)
#dim(PM2.5)

PM2.5 <- PM2.5 %>% arrange(Site.ID, Year)

head(PM2.5)


#write_xlsx(PM2.5, "PM2.5.xlsx")
```


```{r message=FALSE, comment="", include=FALSE}

new.dates <- data.frame(Date = seq.Date(from = as.Date("2009-01-01"), to = as.Date("2012-12-31"), by = "day"))

head(new.dates)
dim(new.dates)
str(PM2.5)

#rm(PM2.5_data)

PM2.5_data <- PM2.5 %>% mutate(
  Date = as.Date(Date, format = "%m/%d/%Y"),
  Site.ID = as.character(Site.ID),
  Site.Name = as.character(Site.Name),
  Season = as.character(Season),
  across(-c(Date, Site.ID, Site.Name, Season), as.numeric)
)


#str(PM2.5_data)
#head(PM2.5_data)

PM2.5_data <- full_join(new.dates, PM2.5_data, by = "Date")

str(PM2.5_data)



```

                                    ### REDUNDANT ###
                                    
```{r eval=FALSE, message=FALSE, comment="", include=FALSE}
RR <- read_excel("C:/Users/Ephra/OneDrive/Documents/New folder/PM2.5_edited.xlsx", sheet = "482011034")

RR <- RR %>% mutate(
  Date = as.Date(Date, format = "%m/%d/%Y"),
  Site.ID = as.character(Site.ID),
  Site.Name = as.character(Site.Name),
  Season = as.character(Season),
  across(-c(Date, Site.ID, Site.Name, Season), as.numeric)
)

RR_date <- data.frame(Date = seq.Date(from = as.Date("2009-01-01"), to = as.Date("2012-12-31"), by = "day"))

RR <- full_join(RR_date, RR)

RR <- RR %>% fill(c(2:5, 7:9), .direction = "down") 
         
RR$Dow <- day(ymd(RR$Date))

RR <- RR %>% mutate( PM2.5.Level = zoo::na.approx(PM2.5.Level, na.rm = FALSE))

colSums(is.na(RR))
str(RR)

```


                               ### PM2.5 INTERPOLATION ###
                              
```{r eval=FALSE, message=FALSE, comment=""}
process_sheet <- function(sheet_name) {
  RR <- read_excel("C:/Users/Ephra/OneDrive/Documents/New folder/PM2.5_edited.xlsx", sheet = sheet_name)
  
  RR <- RR %>%
    mutate(
      Date = as.Date(Date, format = "%m/%d/%Y"),
      Site.ID = as.character(Site.ID),
      Site.Name = as.character(Site.Name),
      Season = as.character(Season),
      across(-c(Date, Site.ID, Site.Name, Season), as.numeric)
    )
  
  RR_date <- data.frame(Date = seq.Date(from = as.Date("2009-01-01"), to = as.Date("2012-12-31"), by = "day"))
  

  RR <- full_join(RR_date, RR)
  
  RR <- RR %>% fill(c(2:5, 7:9), .direction = "down")
  
  RR$Dow <- day(ymd(RR$Date))
  
  RR <- RR %>% mutate(PM2.5.Level = zoo::na.approx(PM2.5.Level, na.rm = FALSE))
  
  return(RR)
}

sheets <- c("481671034", "482010024", "482011034", "482011035", "482011039", "482011050", "483390078", "482011042", "482010416")

results <- lapply(sheets, process_sheet)


combined_results <- bind_rows(results, .id = "Sheet")

colSums(is.na(combined_results))

summary(combined_results$Site.ID == 482011034)

write_xlsx(combined_results, "results.xlsx")


```


                             ### PM2.5 AND WEATHER DATA ANALYSIS ###
                             
```{r eval=FALSE, message=FALSE, comment=""}
EE <- combined_results %>% mutate(PM2.5.Level = zoo::na.approx(PM2.5.Level, na.rm = FALSE))

colSums(is.na(EE))

EE <- EE %>% group_by(Site.ID) %>%
  mutate(Time = row_number()) 

#write_xlsx(EE, "final_results.xlsx")

```


                                 ### WEATHER DATA ANALYSIS ###

```{r message=FALSE, comment="", include=FALSE}
file.paths <- list.files(path = "C:/Users/Ephra/Downloads/", pattern = "Weather_Data.*\\.csv", full.names = TRUE)


data.merge <- lapply(file.paths, read.csv) %>%
  bind_rows()

#str(data.merge)
dim(data.merge)


```


```{r eval=FALSE, message=FALSE, comment="", include=FALSE}

weather <- data.merge %>% select(DATE, STATION, HourlyAltimeterSetting, HourlyRelativeHumidity, HourlyVisibility, HourlyWindDirection, HourlyDewPointTemperature, HourlyPrecipitation, HourlyStationPressure, HourlyWetBulbTemperature, HourlyWindSpeed, HourlyDryBulbTemperature)


Weather <- weather %>% rename(
  Date = DATE, Station = STATION, 
  Alt.meter = HourlyAltimeterSetting, 
  RH = HourlyRelativeHumidity, 
  Visibility = HourlyVisibility, 
  WD = HourlyWindDirection, 
  Dew.Temp = HourlyDewPointTemperature, 
  Precip = HourlyPrecipitation, 
  Pressure = HourlyStationPressure, 
  Wet.Temp = HourlyWetBulbTemperature,
  WS = HourlyWindSpeed,
  Dry.Temp = HourlyDryBulbTemperature)

dim(Weather)
names(Weather)

#str(Weather)

colSums(is.na(Weather))

```


```{r eval=FALSE, message=FALSE, warning=FALSE, comment="", include=FALSE}
Weather <- Weather %>% mutate(
  Date = as.Date(Date),
  Station = as.character(Station),
  across(-c(Date, Station), as.numeric)
)

#Weather$Date <- as.Date(Weather$Date)
#Weather$Station <- as.character(Weather$Station)
#str(Weather)

Weather <- Weather %>% arrange(Date)
  
#write_xlsx(Weather, "weather2.xlsx")
```



               ### GETTING THE MEAN AVERAGES FOR THE DATA ###

```{r eval=FALSE, message=FALSE, comment="", include=FALSE}

weather.new <- Weather %>% group_by(Date) %>%
  summarise(Alt.meter = mean(Alt.meter),
            RH = mean(RH),
            Visibility = mean(Visibility),
            WD = mean(WD),
            Dew.Temp = mean(Dew.Temp),
            Precip = mean(Precip),
            Pressure = mean(Pressure),
            Wet.Temp = mean(Wet.Temp),
            WS = mean(WS),
            Dry.Temp = mean(Dry.Temp)
            
          )

#dim(weather.new)
#str(weather.new)
write_xlsx(weather.new, "WEATHER11.xlsx")

weather.new$Station <- rep("72059400188")
#str(weather.new)
#dim(weather.new)
#View(weather.new)
weather.new$Station <- as.character(weather.new$Station)


Weather <- weather.new %>% select(Date, Station, Alt.meter, RH, Visibility, WD, Dew.Temp, Precip, Pressure, Wet.Temp, WS, Dry.Temp)

Weather$Year <- year(ymd(Weather$Date))

#str(Weather)


```

                        ### WORKING PRECIPITATION ###
                        
```{r eval=FALSE, message=FALSE, comment=""}
data <- read_excel("C:/Users/Ephra/OneDrive/Documents/New folder/weather2.xlsx", sheet = "Sheet2")

dim(data)

data11 <- data %>% select(1, 8) %>%
            group_by(Date) %>% summarize(Precip = mean(Precip))

dim(data11)


data12 <- data %>% select(1, 2, 4, 5, 6, 7, 9, 10, 11, 12)

data13 <- left_join(data12, data11, by = "Date")

dim(data13)

write_xlsx(data13, "datawwwqq.xlsx")





#sum(complete.cases(data13))


```


```{r message=FALSE, comment=""}
data.weather <- read_excel("C:/Users/Ephra/OneDrive/Documents/New folder/Weather.new.xlsx")

Weather <- data.weather %>% mutate(
  Date = as.Date(Date),
  Station = as.character(Station),
  across(-c(Date, Station), as.numeric)
) %>% filter(Date >= as.Date("2009-01-01") & Date <= as.Date("2012-12-31"))

str(Weather)

#head(data.weather)
colSums(is.na(data.weather))




```

                            ### WEATHER DATA INTERPLOATION ###
                            
```{r message=FALSE, comment=""}
library(zoo)

New.weather <- Weather %>% mutate( across(-c(Date, Station), ~ na.approx(., na.rm = FALSE)))

#write_xlsx(New.weather, "weather.now.xlsx")

colSums(is.na(New.weather))

```

```{r eval=FALSE, message=FALSE, comment="", include=FALSE}
ERR <- full_join(EE, New.weather)

final_data <- ERR %>%
  rename(Temp = Dry.Temp, Pressure = Alt.meter) %>%
  select(-c(18))


write_xlsx(final_data, "Final_data.xlsx")
```

                                  
                                  ### THE GAM ###
                                  
```{r eval=FALSE, message=FALSE, comment="", include=FALSE}
KK <- length(unique(final_data$Year))
CC <- length(unique(final_data$Time))
BB <- length(unique(final_data$Season))
AA <- length(unique(final_data$Dow))

names(final_data)
str(final_data)

base.mod <- mgcv::gam(log(PM2.5.Level) ~ s(Time) + s(Dow, k = min(BB, 10)) + s(Year, k = min(KK, 10)) + s(Temp) + s(Precip), data = final_data, method = "REML")

base <- mgcv::gam(log(PM2.5.Level) ~ s(Time) + (Precip) + s(Temp) + s(RH) + s(WS) + s(WD) + s(Pressure) + s(Visibility)+ s(Dow, k = min(BB, 10)) + s(Year, k = min(KK, 10)), data = final, method = "REML")

summary(base)

AIC(base)
BIC(base.mod)
rmse.gam <- sqrt(mean((exp(fitted(base)) - final$PM2.5.Level)^2))
rmse.gam
anova(base)

plot(base.mod, resid = TRUE)

```


                      ### FOR TABLE GENERATION IN PPT ###
                      
```{r eval=FALSE, fig.height=7, fig.width=10, message=FALSE, comment="", include=FALSE}
library(knitr)
#names(Air)
sample_Air <- Air %>% sample_n(7)
sample_Air_Main <- sample_Air %>% select(Date, PM2.5.Level, Site.Name, Longitude, Latitude, Method.Description, Season)

#dim(sample_Air_Main)
#str(sample_Air_Main)
kable(sample_Air_Main, caption = "Table 1.1: Sample Air Quality Data")


```

```{r echo=FALSE, fig.height=7, fig.width=10, message=FALSE, comment=""}
library(knitr)
library(dplyr)
names(New.weather)
sample_Wea <- New.weather %>% sample_n(7)
sample_Wea_Main <- sample_Wea %>% select(Date, Station, Alt.meter, RH, Visibility, WD, Precip, WS, Dry.Temp)

BB <- round(sample_Wea_Main[, -c(1, 2)], 2)
WWW <- cbind(Date = sample_Wea_Main$Date, Station = sample_Wea_Main$Station, BB)

#dim(sample_Air_Main)
#str(sample_Air_Main)
kable(WWW, caption = "Table 1.1: Sample Weather Data")


```




                        ### SUMMARY STATISTICS FOR ALL VARIABLES ###
```{r eval=FALSE, message=FALSE, comment="", include=FALSE}

summary_stats <- final %>%
  summarise(across(c(PM2.5.Level, RH, Temp, Visibility, Pressure, Precip, WS, WD, Time), list(
        Median = median,
    Mean = mean,
    SD = sd,
    Min = min,
    Max = max
  ))) %>%
  pivot_longer(everything(), names_sep = '_', names_to = c('Variable', '.value'))

summary <- data.frame(summary_stats[, 1], round(summary_stats[-c(1)], 2)) 
str(summary_stats)
View(summary)

#kable(summary, caption = "Summary statistics for the variables used for model development")
pander::pander(summary)

